---
title: "Fastest 100 metre times"
author: "Team 09: Matt McCoy and Cyrus Tadjiki"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    theme: journal
    highlight: haddock
    toc: yes
    toc_depth: 3
    toc_float: yes
    keep_md: true
---

```{r setup, include=FALSE}
## This next line sets the default behaviour for all R chunks in the .Rmd document.
## I recomend you take a look here: https://rmarkdown.rstudio.com/authoring_rcodechunks.html
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, error = TRUE, dpi=300)
```

## Background

In class, we practiced webscraping with the Wikipedia page on the [Men's 100 metres world record progression](http://en.wikipedia.org/wiki/Men%27s_100_metres_world_record_progression). For this assignment, we're going to continue with a similar theme, except we won't only be limiting ourselves to world record times. The page that we're scraping will also have a few complications that require extra (or at least different) steps.

Here is the webpage: **[All-time men's best 100m](http://www.alltime-athletics.com/m_100ok.htm)**.

*<b>Note:</b> You are welcome to use the [women's all-time best 100m times](http://www.alltime-athletics.com/w_100ok.htm) if you prefer. However, please be aware that you may (will?) have to adjust some of the specific hints below. It will be become more obvious why once we get to the prediction section of the assignment.*

Now is good time to load any packages that you will be needing, as well as set your preferred plotting theme, etc. 
## Loading Packages
```{r libs, cache=F, message=F}
## Load your packages here, e.g.
if (!require("pacman")) install.packages("pacman")
pacman::p_load(lubridate, janitor, rvest, jsonlite, fredr, listviewer, usethis, sf, tidyverse, data.table, hrbrthemes, lwgeom, rnaturalearth, maps, mapdata, spData, tigris, tidycensus, leaflet, mapview, tmap, tmaptools, xml2, gghighlight, sjPlot, sjmisc, sjlabelled, broom, ggplot2, stargazer, table1)
```


## Things to check
```{r}
# 2.2 I'm not sure if he wants us to exclude spaces or not. I can't figure out how to exclude spaces if so yet. Maybe the string/grep presenter can help? :)
# 3.2 What does the hint "Look for an athlete that has the first name "Kareem" for an illustration of what might go wrong." point us to? 
#
#
#3.4 i'm not sure if im supposed to convert rank to something. some values are weird like 1rA and 2r3 
#
#
#
#6.1 Not sure how nice of a table he wants for our regression 
# Also Overall I wanna double check I'm doing 6.1 correctly
#
#6.2 We should try and have years on the left to answer "what year would you have predicted humanity reaching the current world record time of 9.58 seconds? How does this compare with the year that Usain Bolt actually set it (i.e. 2009). What do you make of this?"

```


## 1) Read in the data

Take a look at the [webpage](http://www.alltime-athletics.com/m_100ok.htm) in your browser. We only want the information contained in the main table at the top (i.e. ignore the rolling starts, manual timing, etc.) Read this table into R and call the resulting object `m100_wp`.

*Hint: In class, we practiced identifying the correct HMTL elements with CSS selectors and SelectorGadget. However, you will almost certainly find it easier / more precise to scrape the table in this example via its XPath. Use your browser's "Inspect" functionality to find and copy over this XPath. Remember to specify the "xpath" argument (instead of the default "css") when pulling this information into R, i.e. `rvest::html_element(xpath = "XPATH_HERE")`.*


## Load Data
```{r}
base_url = read_html("http://www.alltime-athletics.com/m_100ok.htm")


m100_wp =
  base_url %>%
  html_element(xpath = "/html/body/center[3]/pre")
```

## 2) Parse into an R object


### 2.1) Try parsing with `rvest::html_table()`
```{r}
html_table(m100_wp) #doesn't run unfortunately 
```

With the Wikipedia example from class, we were able to parse an HTML table into a data frame simply by using `rvest::html_table()`. What happens if you try that here?

**We are not able to simply convert this object into a html table because we have one very long string that cannot be put into a neat and tidy table. The Xpath we read everything as is a very long string unlike the Wikipedia table**

### 2.2. Try parsing with `rvest::html_text()`
Unfortunately, the HTML object that we've read into R is old-school text. Luckily, we can still extract this text pretty easily into an R string. Do that and name the resulting object `m100_text`. Show me the first 1000 characters.
*Hint: The `head()` function works by elements, not characters. So you'll need some other function to show the first 1000 characters.*
## Parsing object into text
```{r}
m100_wp = html_text(m100_wp) #If not running input must me a vector. Run line ~58 first. 
```

### First 1000 Characters
```{r}
first_1000_characters = substr(m100_wp, start = 1, stop = 1000)  
  

#typeof(m100_wp)

## I'm not sure if he wants us to exclude spaces or not. I can't figure out how to exclude spaces if so yet. 
first_1000_characters 
```


## 3) Convert to a data frame

### 3.1 ) Read as data frame
```{r}
#m100 = read_fwf(m100_wp, skip = 1, n_max = 3436)
m100 = read_table(m100_wp, col_names = FALSE, col_types = NULL, locale = default_locale(),
  na = "NA",
  skip = 1,
  n_max = 3436,
  progress = show_progress(),
  comment = "",
  skip_empty_rows = TRUE)
  
head(m100)
```

At this point, we basically have one loooong string that we need to convert to a data frame. Please do this and assign the resulting object as `m100`. Don't worry about specifying column names yet.

*Hint: You have two (related) options here: Convert either based on whitespace separation of columns, or fixed-width file format.  The `readr` package provides methods for reading both whitespace-separated columns and fixed-width files. See the [reference guide](https://readr.tidyverse.org/reference/index.html#section-read-rectangular-files) guide for more details. If you opt for the latter, please pay particular attention to the option of guessing the fixed column positions based on the the position of empty columns, as well as how many rows are parsed by default to determine the data frame shape. See the Details and Examples sections of the relevant help documentation.*


### 3.2) Inspect and fix (if needed)
```{r}
head(m100, 9)
tail(m100, 5)
summary(m100)
```


What does your resulting `m100` data frame look like? Print the first and last few rows to screen so that I can see it too.

Depending on the functions and arguments that you used in the previous question, you may need to do some additional work to fix the resulting data frame, For example, you should only have 9 columns. If that isn't the case, inspect your dataset and figure out where the problem lies. Fix this issue as best you can, so that you only have nine columns.

*Hint: Look for an athlete that has the first name "Kareem" for an illustration of what might go wrong.*

**My dataset has 9 columns/variables so there is no issue with that. I haven't added column names yet so it has "Usain Bolt" where "athlete" should be and it has his impossible Personal Record 9.58 Seconds where we should have the label "time" etc...**


### 3.3) Assign column names

You should now (hopefully) have nine columns. Assign them the following names: `c("rank", "time", "windspeed", "athlete", "country", "dob", "race_rank", "location", "date")`.

```{r}
colnames(m100)<-c("rank", "time", "windspeed", "athlete", "country", "dob", "race_rank", "location", "date") #This replaced the first row

head(m100)
```


### 3.4 Convert columns to correct classes

Finally, convert your columns to the correct classes. Date columns should be converted to dates, numeric columns should be converted to numeric, etc.

```{r}
sapply(m100, class)
```

```{r}
m100$date <- dmy(m100$date)

#also need to convert dob
m100$dob <- dmy(m100$dob)

sapply(m100, class) #confirm that we successfully converted to class: Date

head(m100) #dates are now year-month-day

```

```{r}
m100$time <- as.numeric(m100$time)

m100$windspeed <- as.numeric(m100$windspeed)

sapply(m100, class) #confirm that we successfully converted to class: numeric

head(m100) 
```


## 4) Plot the data

Plot the data, with the race date on the x-axis and time on the y-axis. Highlight Usain Bolt's times in red.

```{r}
plot1 <- ggplot(m100, aes(x = date, y = time, color = athlete)) + geom_point() + labs(title="Race time by Date", x = "Race Date", y = "Finish time")

plot1

# library(gghighlight) #already loaded

plot1 + gghighlight(athlete == "Usain Bolt") # I get message: "Too many data points, skip labeling"

#ggplot(m100) + aes(x = date, y = time, color = athlete) + geom_point() + labs(title="Race time by Date", x = "Race Date", y = "Finish time") + gghighlight(athlete == "Usain Bolt")
```


## 5) Subset to fastest times per year

It's hard to fit a sensible model to the above data. What might make more sense is to think of 100 metre times as following some kind of (approximately) deterministic process over the years. Subset the data to the fastest time recorded in each year. Call this new data frame `m100_yr` and then repeat the plot above, again highlighting Usain Bolt's times.

```{r}
m100_fastest <- m100 %>% 
  group_by(date) 
#Extract Year from the Date-Time Column
m100_fastest$year <- year(m100_fastest$date)
#now we have the year for each time
m100_fastest2 <- m100_fastest %>% 
    group_by(year) %>% 
    slice(which.min(time))

   

ggplot(m100_fastest2) + aes(x = date, y = time, color = athlete) + geom_point() + labs(title="Race time by Date", x = "Year", y = "Fastest Time") + gghighlight(athlete == "Usain Bolt")

#Because we lost the first row we do not have the fastest time of all time set by Usain Bolt
```


## 6) Modeling and prediction

Imagine that you are back in the year **2005**. You are tasked with fitting a model of year-best 100m times up until that point. Importantly, your model will also be used to predict the trajectory of future 100m times.

### 6.1) Fit a model

Start by fitting a simple regression model of your choice, using data that would have been available to you then (you can include 2005). You are free to use whatever specification you prefer, but please be explicit by writing the model down. (Use dollar signs to demarcate LaTeX equations in R Markdown.) Please also show me the actual regression results in a nicely-formatted table.

*Hint: I'd advise excluding data from before 1975, since we don't have consecutive or consistent records before then.*

```{r}

post74pre05_m100 =
  m100_fastest2 %>%
  filter(date > "1975-01-01", date < "2006-01-01")

#we- estimate the model using a trained sub-sample of the data (1975-2006) and then predict on the full data. 

reg = lm(time ~ date, data = post74pre05_m100) 

plot6 =
  ggplot(data = post74pre05_m100, aes(x= date, y = time)) +
  geom_point() +
  geom_smooth(method = lm) +
  theme_gray()


summary(reg)  
plot6
#tab_model(reg) #Not sure how nice of a table he wants for our regression 

stargazer(reg, type = "text")
```

$\hat{Y} = \hat{\beta_0} + \hat{\beta_1}x_1 + \epsilon$  
$Time_i = \beta_0 + \hat{\beta_1}Date_i + \epsilon$  
$\beta_0 = 1.006*10^{1}$ and $\beta_1 = -2.905*10^{-6}$  

### 6.2) Prediction


Fast forward to the present day. Given your model and the available data in 2005, what year would you have predicted humanity reaching the current world record time of 9.58 seconds? How does this compare with the year that Usain Bolt actually set it (i.e. 2009). What do you make of this?

```{r}
#Now we predict on the full data
predict(reg, newdata = m100_fastest2, interval = "prediction") %>%
  head(5)

# post74pre05_m100 %>%
#  predict(reg, data.frame(date = 2009-08-16))

b0 = 10.06
b1 = -0.000002905
time_exp = b0+b1*(as.numeric(2009-08-16))
time_exp # two things are wrong. 1. There is probably a more advanced way to write this function I can look back at later 2. The expected time shouldn't be above 10 second by 2009. I think that's a bit slow. Could be a rounding error. 
```



### 6.3) Plot your results

Provide a visual depiction of your prediction model. I basically want you to repeat your earlier plot from Question 4, but now with a (95 percent) prediction envelope. The prediction envelope should extend through both the "fitted" (<= 2005) _and_ "predicted" (2006--present) data periods. Make sure that these two periods are clearly demarcated in your plot.

*Hint: geom_smooth() isn't going to help you here because you need to predict out of sample.*

```{r}
## Alternative to predict(): Use augment() to add .fitted and .resid, as well as 
## .conf.low and .conf.high prediction interval variables to the data.
m100_fastest2 = augment(reg, newdata = m100_fastest2, interval = "prediction")
## Show the new variables (all have a "." prefix)
m100_fastest2 %>% select(contains("."), everything()) %>% head()


m100_fastest2 %>%
  ggplot(aes(x = date, y = time, col = rank(date)<="2006-01-01", fill = rank(date)<="2006-01-01")) +
  geom_point(alpha = 0.7) +
  geom_line(aes(y = .fitted)) +
  geom_ribbon(aes(ymin = .lower, ymax = .upper), alpha = 0.3, col = NA) +
  scale_color_discrete(name = "Pre 2006", aesthetics = c("colour", "fill")) +
  labs(
    title = "Predicting Fastest Times by Year"
    )
```


## 7) Map

### 7.1 Static map

Finally, go back to your original `m100` data frame, which contains all the data (i.e. not just the fastest time in each year). I want you to give me a sense of athletic achievement by country (including duplicate observations for the same athlete). Plot a world map, with a colour fill proportional to the number of fastest times that have been contributed by athletes of each country.

*Hint: Use the `sf`-compatible "countries" data frame from the `rnaturalearth` package that I showed you in the spatial lecture. This will come with a column of "iso_a3" country codes that you can match (i.e. join) to the "country" column of the `m100` data frame. There will likely be some mismatches because of inconsistencies in the ISO codes across these data frames. I won't be too strict about this, though you may find the `countrycode::countrycode()` function a helpful alternative.*

```{r}
library(countrycode)
m100$country_code <- countrycode(m100$country,origin = 'wb', destination = 'iso3c')

library(rnaturalearth)
world <- ne_countries(scale = "medium", returnclass = "sf")
class(world)
world

tibble <- m100
tibble2 <- world

joined_tibble <- left_join(tibble, tibble2, 
              by = c("country" = "iso_a3"))



```

```{r}

library(plyr)
country_count <- count(m100$country)
colnames(country_count)<-c("country", "occurance")

new_data <- left_join(country_count, joined_tibble, by = "country")

world_plot = 
  ggplot(new_data) +
  geom_sf(aes(geometry = geometry,fill = occurance), alpha=0.8, col="white") +
  ggtitle("Fastest Times by Country") + scale_fill_viridis_c(name = "occurance")
world_plot


```


### 7.2. Interactive map

A major downside of the above static map is that some powerhouses of world athletics are geographically tiny nations in the Caribbean, which are very hard to see. One way to overcome this is to plot an interactive map using **leaflet** or one of the derivative packages that we discussed in the spatial lecture.


